
---
title: "autotest"
output:
  rmarkdown::html_vignette:
    self_contained: no

  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = TRUE,
  message = TRUE,
  width = 120,
  comment = "#>",
  fig.retina = 2,
  fig.path = "README-"
)
```

<!-- badges: start -->

[![R build
status](https://github.com/ropensci-review-tools/autotest/workflows/R-CMD-check/badge.svg)](https://github.com/ropensci-review-tools/autotest/actions?query=workflow%3AR-CMD-check)
[![codecov](https://codecov.io/gh/ropensci-review-tools/autotest/branch/master/graph/badge.svg)](https://codecov.io/gh/ropensci-review-tools/autotest)
[![Project Status:
Concept](https://www.repostatus.org/badges/latest/concept.svg)](https://www.repostatus.org/#concept)
<!-- badges: end -->

# autotest


Automatic mutation testing of R packages. Mutation in the sense of mutating
inputs (parameters) to function calls, rather than mutation of underlying code
(see, for example, [`mutant`](https://github.com/ropensci/mutant) for that).
`autotest` primarily works by scraping documented examples for all functions,
and mutating the parameters input to those functions.

**This package is very unstable and subject to ongoing development
(`r strftime(Sys.Date(),"%b")`, 2021)**

## Installation

The easiest way to install this package is via the associated
[`r-universe`](https://ropensci-review-tools.r-universe.dev/ui#builds). As
shown there, simply enable the universe with

```{r options, eval = FALSE}
options(repos = c(
    ropenscireviewtools = "https://ropensci-review-tools.r-universe.dev",
    CRAN = "https://cloud.r-project.org"))
```

And then install the usual way with,

```{r install, eval = FALSE}
install.packages("autotest")
```

Alternatively, the package can be installed by running one of the following
lines:

```{r gh-installation, eval = FALSE}
# install.packages("remotes")
remotes::install_git("https://git.sr.ht/~mpadge/autotest")
remotes::install_bitbucket("mpadge/autotest")
remotes::install_gitlab("mpadge/autotest")
remotes::install_github("ropensci-review-tools/autotest")
```

The package can then be loaded the usual way:
```{r load-fakey, eval = FALSE}
library (autotest)
```
```{r load, echo = FALSE, message = FALSE}
devtools::load_all (".", export_all = FALSE)
```

## Usage

The simply way to use the package is

```{r autotest-example, eval = FALSE}
x <- autotest_package ("<package>")
```
The argument to `autotest_package()` can either be the name of an installed
package, or a path to a local directory containing the source for a package.
The result is a `data.frame` of errors, warnings, and other diagnostic messages
issued during package `auotest`-ing. See the [main package
vignette](https://docs.ropensci.org/autotest/articles/autotest.html) for
an introductory tour of the package.

## What is tested?

The package includes a function which lists all tests currently implemented.

```{r autotest_types}
autotest_types ()
```

That functions returns a [`tibble`](https://tibble.tidyverse.org) describing
`r nrow(autotest_types())` unique tests. All `autotest` functions return these
same kinds of objects. The table returned from
[`autotest_types()`](https://docs.ropensci.org/autotest/reference/autotest_types.html)
can be used to selectively switch tests off by setting values in the `test`
column to `FALSE`, as demonstrated below.

Descriptions of each test can be readily extracted from the results of that function:

```{r autotest_type_descs}
a <- autotest_types ()
print (a [, c ("parameter_type", "operation", "content")], n = Inf)
```


## How Does It Work?

The `autotest_package()` function returns by default a list of all tests which
would be conducted on a package, without actually implementing those tests. The
function has a parameter, `test`, with a default value of `FALSE`. Setting
`test = TRUE` then implements all tests, and only returns results from tests
which diverge from expected behaviour, whether unexpected errors, warnings, or
other behaviour. An ideal result is that `autotest_package(., test = TRUE)`
returns nothing (strictly, `NULL`), indicating that all tests passed
successfully.

Tests can also be selectively applied to particular functions through the
parameters `functions`, used to nominate functions to include in tests, or
`exclude`, used to nominate functions to exclude from tests. The following code
illustrates.

```{r stats-var-no-test, fig.show = "hide"}
x <- autotest_package (package = "stats", functions = "var", test = FALSE)
print (x)
```

Testing the `var` function also tests `cor` and `cov`, because the package
works by scraping the documented examples from the associated `.Rd` help file,
and `?var` shows that the help topic is `cor`, and includes the three
functions, `var`, `cor`, and `cov`. That result details the `r nrow (x)` tests
which would be applied to the `var` function from the `stats` package. These
`r nrow (x)` tests yield the following results when actually applied:

```{r stats-var-test}
y <- autotest_package (package = "stats", functions = "var", test = TRUE)
print (y)
```

And only `r nrow (y)` of the original `r nrow (x)` tests produced unexpected
behaviour. There were in fact only three kinds of tests which produced these
`r nrow (y)` results:


```{r unique-operations}
unique (y$operation)
```

The first involves conversion of a vector to a list-column representation (via
`I(as.list(<vec>))`). Relatively few packages accept this kind of input, even
though doing so is relatively straightforward. The following lines demonstrate
how these tests can be switched off when `autotest`-ing a package. The
`autotest_types()` function, used above to extract information on all types of
tests, also accepts a single argument listing the `test_name` entries of any
tests which are to be switched off.

```{r stats-var-test-switch}
types <- autotest_types (notest = "vector_to_list_col")
y <- autotest_package (package = "stats", functions = "var",
                       test = TRUE, test_data = types)
print (y)
```

Those tests are still returned from `autotest_package()`, but with `test =
FALSE` to indicate they were not run, and a `type` of "no_test" rather than the
previous "diagnostic".


## Can `autotest` automatically create tests in my `tests` directory?

Not yet, but that should be possible soon. In the meantime, there are
[`testthat`](https://testthat.r-lib.org) expectations, listed in the [main
package
functions](https://docs.ropensci.org/autotest/reference/index.html),
which enable `autotest` to be used in a package's test suite.


## Prior work

1. The
   [`great-expectations`](https://github.com/great-expectations/great_expectations)
   framework for python, described in [this medium
   article](https://medium.com/@expectgreatdata/down-with-pipeline-debt-introducing-great-expectations-862ddc46782a).
2. [`QuickCheck`](https://hackage.haskell.org/package/QuickCheck) for Haskell
3. [`mutate`](https://github.com/mbj/mutant) for ruby.
