
---
title: "autotest"
output:
  rmarkdown::html_vignette:
    self_contained: no

  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = TRUE,
  message = TRUE,
  width = 120,
  comment = "#>",
  fig.retina = 2,
  fig.path = "README-"
)
```

<!-- badges: start -->

[![R build
status](https://github.com/mpadge/autotest/workflows/R-CMD-check/badge.svg)](https://github.com/mpadge/autotest/actions?query=workflow%3AR-CMD-check)
[![codecov](https://codecov.io/gh/mpadge/autotest/branch/master/graph/badge.svg)](https://codecov.io/gh/mpadge/autotest)
[![Project Status:
Concept](https://www.repostatus.org/badges/latest/concept.svg)](https://www.repostatus.org/#concept)
<!-- badges: end -->

# autotest


Automatic testing of R packages via a simple YAML schema specifying one
or more example workflows for each function of a package. The simplest
workflows involve nominating data sets which may be submitted to a function,
while more complicated workflows may involve multiple data sets, intermediate
pre-processing stages, and other transformations prior to submission to
a nominated function. The following illustrates a typical `yaml` schema, where
items in angle-brackets (`<`, `>`) must be completed. 

```
package: <package_name>
functions:
    - <name of function>:
        - preprocess:
            - "<R code required for pre-processing exlosed in quotation marks>"
            - "<second line of pre-processing code>"
            - "<more code>"
        - parameters:
            - <param_name>: <value>
            - <another_param>: <value>
    - <name of same or different function>::
        - parameters:
            - <param_name>: <value>
```

This `yaml` code should be in the `test` directory of a package -- not in the
`test/testthat` directory. The above template can be generated via the function
`at_yaml_template()`.

## Installation

Not yet on CRAN, so must be installed from remote repository host systems using
any one of the following options:
```{r gh-installation, eval = FALSE}
# install.packages("remotes")
remotes::install_git("https://git.sr.ht/~mpadge/autotest")
remotes::install_bitbucket("mpadge/autotest")
remotes::install_gitlab("mpadge/autotest")
remotes::install_github("mpadge/autotest")
```

The package can then be loaded the usual way:
```{r load-fakey, eval = FALSE}
library (autotest)
```
```{r load, echo = FALSE, message = FALSE}
devtools::load_all (".", export_all = FALSE)
```

## What gets tested?

The package is primarily intended to test packages and functions intended to
process *data*, with data suitable for input into functions within a package
specified in the `autotest.yaml` template. Other types of packages, such as
those which provide access to data, will not be effectively tested by
`autotest`. Standard tests include translation, substitution, and manipulation
of classes and attributes of most standard one- and two-dimensional forms for
representing data in R.

## Example

Current functionality applied to test two functions from the [`SmartEDA`
package](https://github.com/daya6489/SmartEDA) as specified in the following
`yaml`. This also illustrates the use of data transformation steps, in this
case implementing a couple of simple transformations on columns the external
data set, [`ISLR::Carseats`](https://cran.r-project.org/package=ISLR),
including introducing some missing data through replacing values with `NA`.

```{r yaml}
yaml <- c ("package: SmartEDA",
"functions:",
"   - ExpData:",
"       - preprocess:",
"           - 'x <- ISLR::Carseats'",
"           - 'x$Sales <- sqrt (x$Sales)'",
"           - 'x$CompPrice <- as.integer (x$CompPrice)'",
"           - 'x$Sales [ceiling (runif (10) * nrow (x))] <- NA'",
"       - parameters:",
"           - data: x",
"   - ExpData:",
"       - parameters:",
"           - data: ISLR::College",
"   - ExpStat:",
"       - parameters:",
"           - X: ISLR::Carseats$Sales",
"           - Y: ISLR::Carseats$Urban",
"           - valueOfGood: 'Yes'")
```

Both functions `ExpData` and `ExpStat` accept rectangular input, and so can be
tested via the `autotest_rectangular()` function:

```{r autotest, eval = TRUE}
autotest_rectangular (yaml = yaml)
```

All tests work for those functions, so no additional diagnostic output is
generated.
