
---
title: "autotest"
output:
  rmarkdown::html_vignette:
    self_contained: no

  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = TRUE,
  message = TRUE,
  width = 120,
  comment = "#>",
  fig.retina = 2,
  fig.path = "README-"
)
```

<!-- badges: start -->

[![R build
status](https://github.com/ropenscilabs/autotest/workflows/R-CMD-check/badge.svg)](https://github.com/ropenscilabs/autotest/actions?query=workflow%3AR-CMD-check)
[![codecov](https://codecov.io/gh/ropenscilabs/autotest/branch/master/graph/badge.svg)](https://codecov.io/gh/ropenscilabs/autotest)
[![Project Status:
Concept](https://www.repostatus.org/badges/latest/concept.svg)](https://www.repostatus.org/#concept)
<!-- badges: end -->

# autotest


Automatic mutation testing of R packages. Mutation in the sense of mutating
inputs (parameters) to function calls, rather than mutation of underlying code
(see, for example, [`mutant`](https://github.com/ropensci/mutant) for that).
`autotest` primarily works by scraping documented examples for all functions,
and mutating the parameters input to those functions.

**This package is very unstable and subject to ongoing development (Oct 2020)**

## Installation

Not yet on CRAN, so must be installed from remote repository host systems using
any one of the following options:
```{r gh-installation, eval = FALSE}
# install.packages("remotes")
remotes::install_git("https://git.sr.ht/~mpadge/autotest")
remotes::install_bitbucket("mpadge/autotest")
remotes::install_gitlab("mpadge/autotest")
remotes::install_github("ropenscilabs/autotest")
```

The package can then be loaded the usual way:
```{r load-fakey, eval = FALSE}
library (autotest)
```
```{r load, echo = FALSE, message = FALSE}
devtools::load_all (".", export_all = FALSE)
```

## Usage

The simply way to use the package is

```{r autotest-example, eval = FALSE}
x <- autotest_package ("<package>")
```
The argument to `autotest_package()` can either be the name of an installed
package, or a path to a local directory containing the source for a package.
The result is a `data.frame` of errors, warnings, and other diagnostic messages
issued during package `auotest`-ing.


## What gets tested?

The package is primarily intended to test packages and functions intended to
process *data*, with data suitable for input into functions within a package
specified in the `autotest.yaml` template. Other types of packages, such as
those which provide access to data, will not be effectively tested by
`autotest`. Standard tests include translation, substitution, and manipulation
of classes and attributes of most standard one- and two-dimensional forms for
representing data in R.


## Prior work

1. The
   [`great-expectations`](https://github.com/great-expectations/great_expectations)
   framework for python, described in [this medium
   article](https://medium.com/@expectgreatdata/down-with-pipeline-debt-introducing-great-expectations-862ddc46782a).
2. [`QuickCheck`](https://hackage.haskell.org/package/QuickCheck) for Haskell
3. [`mutate`](https://github.com/mbj/mutant) for ruby.
